Linear regression for approximating real and synthetic data

### 2022-10-11 doc/paper.tex
Paper finito

To define the optimal prediction, we use a cost function, and the choice of cost function greatly influences what optimal parameters we get.\footnote{It is here important to keep in mind that the cost function used when fitting the model (training) rarely is the same as the one used for testing our prediction, as described in <somewhere>} An intuitive cost function is the mean squared error (MSE). This function simply computes the average distance between our prediction and the target value, squared:

@@

Secondly, we will perform our linear regression on this new dataset and store the predictions of the test data. Thirdly, we repeat for a given large number of iterations. Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem (REFERENCE!) obtain a normal distribution of these performance metrics. The law of large numbers <reference?> then tells us that the mean sample values of the MSE, bias and variance approach the true value of MSE, bias and variance. This is dependent on our assumption that our data is independently and identically distributed.

@@

Looking at Lasso regression, we find that it has the same capability to reduce overfitting as Ridge, as can be seen in Fig.~\ref{frankelassodifflambdas}. However, looking at Table.~\ref{ols_vs_ridge_vs_lasso_table}, we see that Lasso falls short when finding a good MSE. Intuitively, it does not make sense for Lasso to be worse than OLS, because Lasso with $\lambda = 0$ is equivalent to OLS. But we cannot avoid considering the fact that Lasso does not have an analytical solution, and as such we must use numerical methods to find the optimal parameters. Using scikit's implementation, we have not been able to get the numerical method to converge, which means that we have not found a global minimum.\footnote{Scikit explicitly mentions that using $\lambda = 0$ does not converge well with their implementation} For this reason, Lasso regression has not given satisfying results.

@@

Here we find some limitations of linear regression. We see from Fig.~\ref{real1msetraintest} that the we are able to fit the terrain up to a certain point, gaining an MSE of $0.0142683$ with OLS with polynomial degree 19, our highest polynomial. This is about two orders of magnitude higher than our lowest MSE for the Franke function without noise, which had values on a similar scale.\footnote{The Franke function without noise has a range of about $(0,1.2)$ and our MinMaxed terrain data has been scaled to a range of $(0,1)$. While not exactly the same, these are on the same order of magnitude, and we believe that comparing the MSE scores is reasonable} In fact, our MSE is comparable to the Franke function with a stochastic noise $\epsilon \sim N(0,0.1)$ (Table.~\ref{ols_mse_table_dif_data}), which implies that our terrain is very irregular and difficult to approximate. Looking at the plot for polynomial degree 19 (Fig.~\ref{real1surfacecomp}), we see that our poor MSE is reflected visually in our fit as well, as the surfaces do not look particularly similar. Comparing the best polynomial degree of the best fit with the synthetic data, we see that our best fit is reached much later. This implies that we are not overfitting yet, and indeed the bias variance decomposition confirms this, as we see that variance is essentially inconsequential (Fig.~\ref{realbiasvarianceOLS}). %This can be seen when we increase the polynomial degree to very large amounts, without seeing any decrease in MSE (Fig. NUMBER). Looking at the beta progression for selected betas (Fig. NUMBER), we see that the parameters stabilize, which implies that we are not overfitting, but simply reaching our best possible prediction for polynomial regression. The fact that we are not overfitting is further reinforced by looking at the bias-variance plot in Fig.~\ref{realbiasvarianceOLS}, which shows no increase in variance. 
It is reasonable to expect that our model will not overfit, due to the large increase in data we now use for the fit. Our synthetic data used $20 x 20 = 400$ datapoints, while we now have $320 x 640 = 204 800$ datapoints. This means that for a 19th order polynomial, which will have $210$ features, we will have around a thousand datapoints per feature. If we decrease the number of datapoints availiable considerably, we can see overfitting start to occur (Fig.~\ref{real1mselessdata} in the appendix). So we are not overfitting, but we are not decreasing our MSE either. This is likely because the elevation data has a jagged and complex shape, which cannot be accurately recreated, even with a polynomial of degree $19$. We should remember that the Franke function can be expanded into a series of polynomials, which is why we can expect to get a lower and lower MSE as the polynomial degree increases (so long as we have enough data and it is not noisy). The same is not true for our elevation data, and as such we have little to gain from increasing our model complexity after a certain point. Clearly, linear regression does not offer a very good tradeoff between computational cost and performance for this data.

@@

As our model is not overfitting, we do not expect Ridge to offer much improvement. This proves to be the case, as we see from Fig.~\ref{real1ridgedifflambdas}. Here we see that different values of lambda have very little impact, as there are no parameters that need regularization, and no overfitting to supress. For very large values, we see that we are shrinking the parameters too much, and we see a slight increase in MSE, but overall we do not see much of an effect. We could expect to see that Ridge shows improvement at higher polynomial degrees, when OLS starts to overfit, but due to computational constraints we have not been able to push much further than 19 degrees. Using scikit's GridSearch, we see some surprising results that we assume are due to the fact that we are not overfitting. We see that our best MSE is higher than for OLS, which is surprising given that Ridge with $\lambda = 0$ is equivalent to OLS. However, we must here consider that our lowest OLS MSE was found by bootstrap with 100 iterations, and our grid searched best lambda and MSE was found by crossvalidation with 10 folds. Given that we do not expect large differences in values with either method because we are not overfitting, it is not so unusual that our Ridge MSE is slightly higher.

@@

With lasso regression, we see similar results as with Ridge. Comparing all the three method, we see again that OLS comes out on top. This time, it is not because we pick the polynomial degree that does not overfit, but because we are not overfitting at the highest degree we are using for our prediction. See Table.~\ref{ols_vs_ridge_vs_lasso_table_real}.

\begin{figure} 
\centerline{\includegraphics[width=3.25in]{figure/real1ridgedifflambdas.png}}
\caption{Bias variance tradeoff for ridge regression with different lambdas}
\label{real1ridgedifflambdas}
\end{figure}

\begin{table*}[t]
\caption{Best MSE for OLS, Ridge and Lasso regression on real elevation data}
\begin{center}
\label{ols_vs_ridge_vs_lasso_table_real}
\begin{tabular}{c | l l l}
Regression & MSE & Degree & $\lambda$ \\
\hline
OLS & 0.0142683 & 19 & - \\
Ridge & 0.015433 & 19 & $18.32$ \\
Lasso & 0.0095846 & 10 & $2.5118 \cdot 10^{-6}$ \\
\hline
\end{tabular}
\end{center}
\end{table*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix       %%% starting appendix
\section*{Appendix A: Plots and tables}

@@

\begin{table*}
\caption{This table shows the command to execute, as well as the advanced parameters to change, to reproduce every figure in the report. (More info about the scripts and their parameters can be found in the README)}
\begin{center}
\label{allparamstable}
\begin{tabular}{c | l l l l}
Figure & Shell command (leading \texttt{python3} omitted) & Bootstraps & K-folds & $\lambda$ \\
\hline
\ref{frankenonoise} & \texttt{task\_b.py --noise 0 -n 25} & - & - & - \\
\ref{frankenonoisemse} & \texttt{task\_c.py --noise 0 -n 24}  & 100 & - & - \\
\ref{frankenoisemse} & \texttt{task\_c.py -n 24}  & 100 & - & - \\
\ref{frankiebiasvariance} & \texttt{task\_c.py -n 24 --step 0.1}  & 100 & - & - \\
\ref{frankiedifflambdas} & \texttt{task\_e\_biasvariance.py -n 24}  & 100 & - & - \\
\ref{real1ridgedifflambdas} & \texttt{task\_e\_biasvariance.py --file ../data/small\_*\_1.tif -n 24}  & 100 & - & - \\
\ref{frankeridgebetas} & \texttt{task\_e\_betas.py -n 24}  & 100 & - & - \\
\ref{frankenoisybeta} & \texttt{task\_c.py -n 24}  & 100 & - & - \\
\ref{frankienoisecomparison} & \texttt{noise\_plot.py} & 100 & - & - \\
\ref{frankeridgenoisecomparison} & \texttt{noise\_plot.py} \footnotesize{(\it Model has been changed to ridge with lambda=$10^{-8}$)} & 100 & - & - \\
\ref{frankiehigherdatapoints} & \texttt{task\_c.py --noise 0 -n 30 --step 0.01} & 100 & - & - \\
\ref{real1surfacecomp} & \texttt{task\_b.py --file ../data/small\_*\_1.tif -n 19} & - & - & - \\
\ref{real1mselessdata} & \texttt{task\_c.py --file ../data/tiny\_SRTM\_data\_Norway\_1.tif -n 19} & 100 & - & - \\
\ref{real1msetraintest} & \texttt{task\_c.py --file ../data/small\_*\_1.tif -n 19} & 100 & - & - \\
\ref{realbiasvarianceOLS} & \texttt{task\_c.py --file ../data/small\_*\_1.tif -n 19} & 100 & - & - \\
\hline
\end{tabular}
\end{center}
\end{table*}

@@

@@

@@

>>>>

To define the optimal prediction, we use a cost function, and the choice of cost function greatly influences what optimal parameters we get.\footnote{It is here important to keep in mind that the cost function used when fitting the model (training) rarely is the same as the one used for testing our prediction} An intuitive cost function is the mean squared error (MSE). This function simply computes the average distance between our prediction and the target value, squared:

@@

Secondly, we will perform our linear regression on this new dataset and store the predictions of the test data. Thirdly, we repeat for a given large number of iterations. Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem \cite{CLT} obtain a normal distribution of these performance metrics. The law of large numbers \cite{LLN} then tells us that the mean sample values of the MSE, bias and variance approach the true value of MSE, bias and variance. This is dependent on our assumption that our data is independently and identically distributed.

@@

Looking at Lasso regression, we find that it has the same capability to reduce overfitting as Ridge, as can be seen in Fig.~\ref{frankelassodifflambdas}. However, looking at Table.~\ref{ols_vs_ridge_vs_lasso_table}, we see that Lasso falls short when finding a good MSE. Intuitively, it does not make sense for Lasso to be worse than OLS, because Lasso with $\lambda = 0$ is equivalent to OLS. But we cannot avoid considering the fact that Lasso does not have an analytical solution, and as such we must use numerical methods to find the optimal parameters. Using scikit's implementation, we have not been able to get the numerical method to converge, which means that we have not found a global minimum.\footnote{Scikit explicitly mentions that using $\lambda = 0$ does not converge well with their implementation \cite{lasso}} For this reason, Lasso regression has not given satisfying results.

@@

Here we find some limitations of linear regression. We see from Fig.~\ref{real1msetraintest} that the we are able to fit the terrain up to a certain point, gaining an MSE of $0.0142683$ with OLS with polynomial degree 19, our highest polynomial. This is about two orders of magnitude higher than our lowest MSE for the Franke function without noise, which had values on a similar scale.\footnote{The Franke function without noise has a range of about $(0,1.2)$ and our MinMaxed terrain data has been scaled to a range of $(0,1)$. While not exactly the same, these are on the same order of magnitude, and we believe that comparing the MSE scores is reasonable} In fact, our MSE is comparable to the Franke function with a stochastic noise $\epsilon \sim N(0,0.1)$ (Table.~\ref{ols_mse_table_dif_data}), which implies that our terrain is very irregular and difficult to approximate. Looking at the plot for polynomial degree 19 (Fig.~\ref{real1surfacecomp}), we see that our poor MSE is reflected visually in our fit as well, as the surfaces do not look particularly similar. Comparing the best polynomial degree of the best fit with the synthetic data, we see that our best fit is reached much later. This implies that we are not overfitting yet, and indeed the bias variance decomposition confirms this, as we see that variance is essentially inconsequential (Fig.~\ref{realbiasvarianceOLS}). The fact that we are not overfitting can be seen when we increase the polynomial degree to very large amounts, seeing only a very modest decrease in MSE to $0.01381$ (Fig.~\ref{realOLSmse40} in the appendix). Comparing this to tenfold decrease in MSE when increasing the polynomial degree for the Franke function, we see a clear difference. Looking at the beta progression for selected betas (Fig.\ref{realOLSbetaprog40} in the appendix), we see that the parameters stabilize, which again implies that we are not overfitting, but simply reaching our best possible prediction for polynomial regression. The fact that we are not overfitting is further reinforced by looking at the bias-variance plot in Fig.~\ref{realbiasvarianceOLS}, which shows no increase in variance. It is reasonable to expect that our model will not overfit, due to the large increase in data we now use for the fit. Our synthetic data used $20 x 20 = 400$ datapoints, while we now have $320 x 640 = 204 800$ datapoints. This means that for a 19th order polynomial, which will have $210$ features, we will have around a thousand datapoints per feature. If we decrease the number of datapoints availiable considerably, we can see overfitting start to occur (Fig.~\ref{real1mselessdata} in the appendix). So we are not overfitting, but we are not decreasing our MSE either. This is likely because the elevation data has a jagged and complex shape, which cannot be accurately recreated, even with a polynomial of degree $19$. We should remember that the Franke function can be expanded into a series of polynomials, which is why we can expect to get a lower and lower MSE as the polynomial degree increases (so long as we have enough data and it is not noisy). The same is not true for our elevation data, and as such we have little to gain from increasing our model complexity after a certain point. Clearly, linear regression does not offer a very good tradeoff between computational cost and performance for this data.

@@

As our model is not overfitting, we do not expect Ridge to offer much improvement. This proves to be the case, as we see from Fig.~\ref{real1ridgedifflambdas} in the appendix. Here we see that different values of lambda have very little impact, as there are no parameters that need regularization, and no overfitting to supress. For very large values, we see that we are shrinking the parameters too much, and we see a slight increase in MSE, but overall we do not see much of an effect. We could expect to see that Ridge shows improvement at higher polynomial degrees, when OLS starts to overfit, but due to computational constraints we have not been able to push much further than 19 degrees. Using scikit's GridSearch, we see some surprising results that we assume are due to the fact that we are not overfitting. We see that our best MSE is higher than for OLS, which is surprising given that Ridge with $\lambda = 0$ is equivalent to OLS. However, we must here consider that our lowest OLS MSE was found by bootstrap with 100 iterations, and our grid searched best lambda and MSE was found by crossvalidation with 10 folds. Given that we do not expect large differences in values with either method because we are not overfitting, it is not so unusual that our Ridge MSE is slightly higher.

@@

With lasso regression, we see similar results as with Ridge. Fig.~\ref{real1lassobiasvariance} in the appendix shows little change, except for when our $\lambda$ is too high and we force all parameters to $0$. Comparing lasso to the two other methods for the real dataset has proven to be difficult, as the numerical methods for finding the best parameters are very time consuming. We have not been able to gridsearch up to polynomial degree 19, but up to degree 9 we found the best lambda to be $2.154^{-7}$ for polynomial degree 9, giving an MSE of $0.01685$.

Comparing all three methods up to degree 9 for different lambdas (using the same lambda for Lasso and Ridge, just to get an overview) in Fig.~\ref{real1msecomparison} in the appendix, we see the same trend as we have been describing. Regularization offers little reward, which is to be expected when we are not overfitting.

\begin{table*}[t]
\caption{Best MSE for OLS and Ridge regression on real elevation data}
\begin{center}
\label{ols_vs_ridge_vs_lasso_table_real}
\begin{tabular}{c | l l l}
Regression & MSE & Degree & $\lambda$ \\
\hline
OLS & 0.0142683 & 19 & - \\
Ridge & 0.015433 & 19 & $18.32$ \\
\hline
\end{tabular}
\end{center}
\end{table*}

@@

\section*{Appendix A: Plots and tables}

@@

\begin{table*}
\caption{This table shows the command to execute to reproduce every figure in the report. (More info about the scripts and their parameters can be found in the README)}
\begin{center}
\label{allparamstable}
\begin{tabular}{c | l l l}
Figure & Shell command (leading \texttt{python3} omitted) \\
\hline
\ref{frankenonoise} & \texttt{task\_b.py --noise 0 -n 25}\\
\ref{frankenonoisemse} & \texttt{task\_c.py --noise 0 -n 24}\\
\ref{frankenoisemse} & \texttt{task\_c.py -n 24}\\
\ref{frankiebiasvariance} & \texttt{task\_c.py -n 24 --step 0.1}\\
\ref{frankiedifflambdas} & \texttt{task\_e\_biasvariance.py -n 24}\\
\ref{real1ridgedifflambdas} & \texttt{task\_e\_biasvariance.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 24}\\
\ref{real1msecomparison} & \texttt{task\_f\_biasvariance.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 9}\\
\ref{frankeridgebetas} & \texttt{task\_e\_betas.py -n 24}\\
\ref{frankenoisybeta} & \texttt{task\_c.py -n 24}\\
\ref{frankienoisecomparison} & \texttt{noise\_plot.py}\\
\ref{frankeridgenoisecomparison} & \texttt{noise\_plot.py} \footnotesize{(\it Model has been changed to ridge with lambda=$10^{-8}$)}\\
\ref{frankiehigherdatapoints} & \texttt{task\_c.py --noise 0 -n 30 --step 0.01}\\
\ref{real1surfacecomp} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 19}\\
\ref{real1mselessdata} & \texttt{task\_c.py --file ../data/tiny\_SRTM\_data\_Norway\_1.tif -n 19}\\
\ref{real1msetraintest} & \texttt{task\_c.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 19}\\
\ref{realbiasvarianceOLS} & \texttt{task\_c.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 19}\\
\ref{realOLSmse40} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 40}\\
\ref{realOLSbetaprog40} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 40}\\
\ref{real1msecomparison} & \texttt{task\_f\_msecomparison.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 9}\\
\hline
\end{tabular}
\end{center}
\end{table*}

@@

\begin{figure*}
\centerline{\includegraphics[width=3.25in]{figure/realOLSmse40.png}}
\caption{Test and train MSE on our terrain, this time up to degree 40. We see only a slight decrease in MSE, down to $0.01381$, even though we double the amount of degrees. This has not been bootstrapped due to computational limitations}
\label{realOLSmse40}
\end{figure*}

\begin{figure*}
\centerline{\includegraphics[width=3.25in]{figure/realOLSbetaprog40.png}}
\caption{Beta progression up to polynomial degree 40 for real terrain data, showing that the are relatively constant, implying that we are not overfitting}
\label{realOLSbetaprog40}
\end{figure*}

@@

\begin{figure*}
\centerline{\includegraphics[width=3.25in]{figure/real1msecomparison.png}}
\caption{Comparison between OLS, Ridge and Lasso for different values of lambda, for real terrain data}
\label{real1msecomparison}
\end{figure*}

@@

\begin{figure} 
\centerline{\includegraphics[width=3.25in]{figure/real1ridgedifflambdas.png}}
\caption{Bias variance tradeoff for ridge regression with different lambdas, for the real dataset}
\label{real1ridgedifflambdas}
\end{figure}

\begin{figure} 
\centerline{\includegraphics[width=3.25in]{figure/real1lassobiasvariance.png}}
\caption{Bias variance tradeoff for lasso regression with different lambdas, for the real dataset}
\label{real1lassobiasvariance}
\end{figure}
